---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->



```{r, include=FALSE}
options("width"=110)
tmp <- packageDescription( basename(getwd()) )
```

```{r, results='asis', echo=FALSE}
cat("#", tmp$Title)
```


**Status**



<a href="https://travis-ci.org/petermeissner/wikipediatrend">
<img src="https://api.travis-ci.org/petermeissner/wikipediatrend.svg?branch=master">
<a/>
<a href="https://codecov.io/gh/petermeissner/wikipediatrend">
  <img src="https://codecov.io/gh/petermeissner/wikipediatrend/branch/master/graph/badge.svg" alt="Codecov" />
</a>
<a href="https://cran.r-project.org/package=wikipediatrend">
<img src="http://www.r-pkg.org/badges/version/wikipediatrend">
</a>
<img src="http://cranlogs.r-pkg.org/badges/grand-total/wikipediatrend">
<img src="http://cranlogs.r-pkg.org/badges/wikipediatrend">






```{r, include=FALSE}
filelist.R   <- list.files("R", recursive = TRUE, pattern="\\.R$", ignore.case = TRUE, full.names = TRUE)
filelist.tests   <- list.files("tests", recursive = TRUE, pattern="\\.R$", ignore.case = TRUE, full.names = TRUE)
filelist.cpp <- list.files("src", recursive = TRUE, pattern="\\.cpp$", ignore.case = TRUE, full.names = TRUE)
lines.R      <- unlist(lapply(filelist.R, readLines))
lines.tests  <- unlist(lapply(filelist.tests, readLines))
lines.cpp    <- unlist(lapply(filelist.cpp, readLines))
length.R     <- length(grep("(^\\s*$)|(^\\s*#)|(^\\s*//)", lines.R,  value = TRUE, invert = TRUE))
length.tests <- length(grep("(^\\s*$)|(^\\s*#)|(^\\s*//)", lines.tests,  value = TRUE, invert = TRUE))
length.cpp   <- length(grep("(^\\s*$)|(^\\s*#)|(^\\s*//)", lines.cpp,  value = TRUE, invert = TRUE))
```


*lines of R code:* `r length.R`, *lines of test code:* `r length.tests`



**Version**

```{r, include=FALSE}
source_files <- 
  grep(
    "/R/|/src/|/tests/",
    list.files(recursive = TRUE, full.names = TRUE), 
    value = TRUE
  )
last_change <- 
  as.character(
    format(max(file.info(source_files)$mtime), tz="UTC")
  )
```


```{r, results='asis', echo=FALSE}
cat(tmp$Version, "(",last_change,")")
```

**Description**


```{r, results='asis', echo=FALSE}
#cat(tmp$Description)
```


**License**

```{r, results='asis', echo=FALSE}
cat(tmp$License, "<br>")
cat(tmp$Author)
```


**Credits**

- Parts of the package's code have been shamelessly copied and modified from R base package written by R core team. This concerns the `wp_date()` generic and its methods and is detailed in the help files. 




**Citation**


```{r, results='asis', eval=FALSE}
citation("wikipediatrend")
```

```{r, results='asis', echo=FALSE}
print(citation("wikipediatrend"), style = "text")
```

**BibTex for citing**

```{r, eval=FALSE}
toBibtex(citation("wikipediatrend"))
```



**Installation**

Stable version from CRAN:

```{r, eval=FALSE}
install.packages("wikipediatrend")
```


Latest development version from Github:

```{r, eval=FALSE}
devtools::install_github("petermeissner/wikipediatrend")
```


**Usage**

*starting up ...*
```{r}
library(wikipediatrend)
```


*getting some data ...*
```{r}
trend_data <- 
  wp_trend(
    page = c("Der_Spiegel", "Die_Zeit"), 
    lang = c("de", "en"), 
    from = "2007-01-01",
    to   = Sys.Date()
  )
```

*having a look ...*

```{r}
trend_data
```


*having another look ...*

```{r, fig.width=10}
plot(
  trend_data[trend_data$views < 2500, ]
)
```




