---
title: "wikipediatrend"
output:
  html_document:
    keep_md: yes
    self_contained: yes
---

# Author

Peter Meißner



# Last Update

`r Sys.Date()`


```{r,include=FALSE}
library(wikipediatrend)
```


# Status (current version on Github)

<table>
<tr>
<td> 
Ubuntu build 
</td>
<td> 
<a href="https://travis-ci.org/petermeissner/wikipediatrend">
<img src="https://api.travis-ci.org/petermeissner/wikipediatrend.svg?branch=master">
</a>
</td>
</tr>
<tr>
<td> 
Windows build
</td>
<td> 
<a href="https://ci.appveyor.com/project/petermeissner/wikipediatrend">
<img src="http://ci.appveyor.com/api/projects/status/github/petermeissner/wikipediatrend">
</a>
</td>
</tr>
<tr>
<td>
Version on CRAN  
</td> 
<td>
<a href="https://cran.r-project.org/web/packages/wikipediatrend/index.html">
<img src="http://www.r-pkg.org/badges/version/wikipediatrend">
</a>
</td>
</tr>
<tr>
<td>
Version on Github
</td> 
<td>
`r sessionInfo()$otherPkgs$wikipediatrend$Version`
</td>
</tr>
<tr>
<td>
Downloads from <a href='http://cran.rstudio.com/'>CRAN.RStudio</a>&nbsp;&nbsp;&nbsp;
</td>
<td>
<img src="http://cranlogs.r-pkg.org/badges/grand-total/wikipediatrend">
<img src="http://cranlogs.r-pkg.org/badges/wikipediatrend">
</td>
</tr>

</table>





# Purpose

The wikipediatrend package is designed to make Wikipedia page access statistics data availible in R in a most convenient way. 

*Consequently the package provides* 

- daily page views as data frames 
- page views for user set time spans
- page views for multiple articles in one function call
- page views for articles in different language domains
- a function to check article titles in other country domains
- background caching of results to minimize function execution time as well as server burdens






# Installation 

A stable version of the package can be found on CRAN and installed via ...
```{r, eval=FALSE}
install.packages("wikipediatrend")
```

... while the current developement version can be retrieved by using `install_github()` from the devtools package ... 

```{r, eval=FALSE}
devtools::install_github("petermeissner/wikipediatrend")
```

After loading the package several functions are available.

```{r, error=TRUE, message=FALSE}
library(wikipediatrend)
```




# Usage


The workhorse of the package is the `wp_trend()` function:

```{r, cache=TRUE, error=TRUE}
wp <- wp_trend(page = c("Fever","Fieber"), 
               from = "2013-08-01", 
               to   = prev_month_end(), 
               lang = c("en","de"))
```

The function's return is a data frame with six variables *date*, *count*, *project*, *title*, *rank*, *month* paralleling the data provided by the stats.grok.se server:

```{r, cache=TRUE, error=TRUE}
head(wp)
```

```{r, cache=TRUE, error=TRUE, message=FALSE}
library(ggplot2)

ggplot(wp, aes(date, count, group=page, color = page)) + 
  geom_point() +
  geom_smooth(method="lm", formula = y ~ poly(x, 20), size=1.5) +
  theme_bw()
```




# Vignette

*For a more detailed usage have a look at the vignette accompanying the package. `vignette("using-wikipediatrend", package="wikipediatrend")`*

... or GoTo [CRAN](http://cran.r-project.org/web/packages/wikipediatrend/index.html) or build it from scratch from [Github](https://raw.githubusercontent.com/petermeissner/wikipediatrend/master/vignettes/using-wikipediatrend.Rmd).



# Some examples for using page view statistics


- Wilkerson, Bill (2015): *Post-Republican debate on Wikipedia follow-up: before and after public interest in the candidates* http://www.wrwilkerson.com/ . 
http://www.wrwilkerson.com/blog/2015/8/15/post-republican-debate-on-wikipedia-follow-up-before-and-after-public-interest-in-the-candidates

- Mellon, Jonathan (2014) *Internet Search Data and Issue Salience: The Properties of Google Trends as a Measure of Issue Salience* Journal of Elections, Public Opinion and Parties 24(1):45-72.
http://www.tandfonline.com/doi/abs/10.1080/17457289.2013.846346 

- Ripberger, Joseph T. (2011): *Capturing curiosity: using Internet search trends to measure public attentiveness*. Policy Studies Journal 39(2):239-259.
http://onlinelibrary.wiley.com/doi/10.1111/j.1541-0072.2011.00406.x/full

- Yla Tausczik, Kate Faasse, James W. Pennebaker, Keith J. Petrie (2012): *Public Anxiety and Information Seeking Following the H1N1 Outbreak: Blogs, Newspaper Articles, and Wikipedia Visits*. Health Communication, Vol. 27, Iss. 2.
 http://www.tandfonline.com/doi/pdf/10.1080/10410236.2011.571759

- Taha Yasseri and Jonathan Bright (2015): *Predicting elections from online information flows: towards theoretically informed models*. http://arxiv.org/abs/1505.01818

 
*(I missed your application? Let me know and I put it here)*



# Thanks 

Fernando Reis, Eryk Walczak, Simon Munzert, Kristin Lindemann





# Credits

- Parts of the package's code have been shamelessly copied and modified from R base package written by R core team. This concerns the `wp_date()` generic and its methods and is detailed in the help files. 



# Meta ([cranlogs](https://github.com/metacran/cranlogs)) wikipediatrend

```{r, fig.width=10, echo=FALSE, fig.retina=TRUE, message=FALSE}
library(cranlogs)
library(jsonlite)
library(stringr)

wptrendversions <- 
  grep(
    "version \\d{1,3}.\\d{1,3}.\\d{1,3}.*//", 
    readLines("NEWS"),
    value = TRUE
  )
wptrendversions <- 
  data.frame(
    version = str_extract(wptrendversions, "\\d{1,3}.\\d{1,3}.\\d{1,3}"),
    date = str_extract(wptrendversions, "\\d{4}-\\d{2}-\\d{2}")
  )

rversions <- fromJSON("http://rversions.r-pkg.org/r-versions")

downloads <- 
  cran_downloads(
    package = "wikipediatrend", 
    from = "2014-09-01", 
    to = Sys.Date()
  )[,1:2]

downloads_cum <- 
  data.frame(
    date = downloads$date,
    sum = round(cumsum(downloads$count)/1000,1)
  )
downloads_cum <- 
  downloads_cum[
    !duplicated(downloads_cum$sum) &
    downloads_cum$sum %% 0.5 == 0 , ]
downloads_cum$date[1] <- "2014-11-07"

plot(downloads, type="l", ylim = c(-40,80))
text(as.Date(rversions$date),-17,labels = rversions$version, col = "red", srt=0, cex=0.7)
#segments(x0=as.Date(rversions$date),x1=as.Date(rversions$date),y0=0,y1=10,col="red", lwd=2)
text(as.Date(wptrendversions$date),-25,labels = wptrendversions$version, col = "steelblue", srt=0, cex=0.7)
#segments(x0=as.Date(wptrendversions$date),x1=as.Date(wptrendversions$date),y0=0,y1=10,col="steelblue",lwd = 2)
text(downloads_cum$date, -34, downloads_cum$sum, cex=0.7, col = "darkgreen")
text(as.Date("2014-08-15"),-25,"wp versions", col="steelblue", cex=0.7,pos=4)
text(as.Date("2014-08-15"),-17,"R versions", col="red", cex=0.7,pos=4)
text(as.Date("2014-08-15"),-34,"K downloads", col="darkgreen", cex=0.7,pos=4)
abline(h=c(0),col="black")
abline(h=c(-40,-10),col="grey")
lines(
  loess.smooth(downloads$date, downloads$count, span = 1/5, degree = 2,
    family = c("gaussian"), evaluation = 50),
  col="darkgreen", lwd=2)
```



