---
title: Introducing Wikipediatrend -- Easy Analyses of Public Attention, Anxiety and
  Information Seeking
author: "Peter Meißner"
output:
  html_document:
    keep_md: yes
    self_contained: no
---

`r Sys.Date()`

*Current version auto-build status:*

<img src="https://api.travis-ci.org/petermeissner/wikipediatrend.svg?branch=master"></img>

<img src="http://cranlogs.r-pkg.org/badges/wikipediatrend"></img>

## Introduction

Besides the well know information provided at Wikipedia -- the articles. Wikipedia provides a lot of valuable meta data, too. 
One type of information are page access statistics -- e.g. http://stats.grok.se/en/201409/Peter_Principle. Another type are the info pages -- e.g. https://en.wikipedia.org/w/index.php?title=Peter_Principle&action=info. While the latter falls into the jurisdiction of the [MediaWiki](http://cran.r-project.org/web/packages/WikipediR/index.html)-package , 
the wikipediatrend package allows easy access to Wikipedia page view statistics. 


## Stats.grok.se API and the wikipediatrend package

The server `http://stats.grok.se` retrieves Wikipedia page access statistics on a daily basis and makes them available to the puplic. Page views are presented per article and language for a whole month for each single day. In addition to the HTML version with that presents the data as bar chart ... 


http://stats.grok.se/en/201409/Peter_Principle

... daily page views are availible in JSON fomrat as well ...

http://stats.grok.se/json/en/201409/Peter_Principle



The wikipeditrend package is designed around the idea to make these data availible in R in a most convenient way. 

*Consequently the package provides* 

- daily page views as data frames 
- page views for user set time spans
- page views for multiple articles in one function call
- page views for articles in different language domains 
- background caching of results minimize function execution time as well as server burdens



## Using the package

### Installation 

A stable version of the package can be found on CRAN and installed via ...
```{r, eval=FALSE}
install.packages("wikipediatrend")
```

... while the current developement version can be retrieved by using `install_github()` from the devtools package ... 

```{r, eval=FALSE}
devtools::install_github("petermeissner/wikipediatrend")
```

After loading the package several functions are available.

```{r, error=TRUE, message=FALSE}
library(wikipediatrend)
```

The workhorse of the package is the `wp_trend()` function with several arguments:

- **page**        []: <br>
... here goes the name or names of the page or pages one might want to get page views from, e.g.: `c("cat","dog")`)

- **from**        [ `Sys.Date()-30` ]: <br>
... starting date of the timespan to be considered

- **to**          [ `Sys.Date()` ]: <br>
... end date of the timespan to be considered

- **lang**        [ `"en"` ]: <br>
... language of the page (might also be a vector of languages, e.g.: `c("de","en")`)

- **file**        [ `.wp_trend_cache` ]: <br>
... file to be used as cache / data storage (By default all data retreived will be stored in a temporary file. If a file name is provided, date will be stored there as CSV. Data will always be appended to an already existing data storage file.)


<!-- thrown out since version 0.3.xxx
  - **friendly**    [ `F` ]: <br>
  ... should `wp_trend()` try minimize workload on behalf of `stats.grok.se`
  - **requestFrom** [ `"anonymous"` ]: <br>
  ... do you care to identify yourself towards `stats.grok.se`
  - **userAgent** [ `FALSE` ] <br>
  ... do you care to send information on your plattform, R version and the package used to make server requests
-->

Let's have a first run using the defaults and telling the function to get page views for the article on the peter principle:

```{r, cache=TRUE, error=TRUE}
peter_principle <- wp_trend("peter_principle")
```

The function's return is a data frame with six variables *date*, *count*, *project*, *title*, *rank*, *month* paralleling the data provided by the stats.grok.se server:

```{r, cache=TRUE, error=TRUE}
head(peter_principle)
```


We can use this information to visualize the page view trend. Using `wp_wday()` we can furthermore discriminate weekdays <span style="color:red">(red)</span> from weekends <span style="color:black">(black)</span>. 

```{r, cache=TRUE, error=TRUE, message=FALSE}
library(ggplot2)

weekend <- ifelse(wp_wday(peter_principle$date) > 5, "Sa-Su", "Mo-Fr")

ggplot(peter_principle, aes(date, count)) + 
  geom_line() + 
  geom_point(size=4, aes(date, count, color=weekend)) +
  theme_bw()
```

Looking at the graph we can conclude that the *Peter Principle* as a work related phenomenon obviously is something that is most pressing on workdays -- or maybe people in general just tend to use their computers less on weekends -- we might never know for certain. 







## Being friendly

One of the most important features of the package is its `friendly` option. On the one hand, it saves us time when making subsequent requests of the same page because less pages have to be loaded. On the other hand, it serves to minimize workload on behalf of the `stats.grok.se`-server that kindly provides the information we are using. 

The option can be set to different values: 

- **FALSE**, the default, deactivates `wp_trend()`'s friendly behavior altogether
- **TRUE**, activates `wp_trend()`'s friendly behavior and retreieved access statistics are stored on disk in CSV format via `write.csv()`
- **1** is the same as **TRUE**
- **2**, is the same as **TRUE** but storage takes place via `write.csv2()`

Let's try it out by making two subsequent requests to get access statistics for for information on ISIS. 

```{r, results='hide', error=FALSE, message=FALSE, warning=FALSE}
file.remove("../isil.csv")
```

While for the first request the server has to provide information many times, the second request only asks for those months for which we do not have complete data already. Furthermore, `wp_trend()` informs us that the data has been stored in a CSV-file.



```{r, cache=FALSE, error=TRUE}
isil <- wp_trend("Islamic_State_of_Iraq_and_the_Levant", from="2013-01-01", file="../isil.csv")
```

The second request uses this previous saved information to minimize traffic and function execution time. If it downloads new data, it updates the data already stored on disk.


```{r, cache=FALSE, error=TRUE}
isil <- wp_trend("Islamic_State_of_Iraq_and_the_Levant", from="2012-12-01", file="../isil.csv")
```

Last but not least, let's have a look at the data ... 

```{r, cache=FALSE, error=TRUE}
plot( isil[,1:2], 
      ylim=c(0, max(isil$count)),
      main="ISIL's Wikipedia Attention",
      ylab="views per day", xlab="time",
      type="l")
```

... revealing what most might have already suspected: ISIL is quite a new phenomenon. 


## So what? 

### Cats

First of all we are now able to study cats:

```{r, cache=FALSE, error=TRUE}
catdog <- wp_trend( c("cat","dog"), 
                    from = "2013-01-01", 
                    to   = "2014-03-01", 
                    file = "../wp_trend_cache.csv")
plot( cats[,1:2], 
      col="black",
      ylim=c(0, max(cats$count)),
      main="Cats' Wikipedia Attention",
      ylab="views per day", xlab="time", type="h")
soo_2012_13 <- wp_year(cats$date)== 2012 | wp_year(cats$date)== 2013 
cats_model  <- lm(count ~ date + date^2 + date^3 + soo_2012_13 ,data=cats)
cats_smooth <- data.frame(date=cats$date, count_smooth=predict(cats_model))
lines(cats_smooth,col=rgb(1,0,0,0.5),lwd=5)
```

... and triumphantly can conclude: 

**Cats' popularity is in decline overal and more so cats are soooo old fashioned 2012 and 2013.**.

### Ebola

Or we can study how the desire to inform oneself about Ebola varies over time:

```{r, cache=FALSE, error=TRUE}
ebola_en <- wp_trend("Ebola", from="2008-01-01", file="../wp_trend_cache.csv")
plot( ebola_en[,1:2], 
      ylim=c(0, max(ebola_en$count)),
      main="Ebola's Wikipedia Attention",
      ylab="views per day", xlab="time",
      type="l")
lines(ebola_en)
```

Which unsurprisingly peaks in 2014 with the Ebola outbreak in Western Africa. 

Using the language option we can also study if media attentions differ between languages / cultures by comparing the attention patterns for the English Wikipedia with those for the German Wikipedia:


```{r, cache=FALSE, error=TRUE}
ebola_de <- wp_trend("Ebola", lang="de", from="2008-01-01", file="../wp_trend_cache.csv")
```

```{r, cache=FALSE, error=TRUE}
plot( ebola_en[,1:2], 
      ylim=c(0, max(ebola_en$count)),
      main="Ebola's Wikipedia Attention",
      ylab="views per day", xlab="time",
      type="n")
lines(ebola_en[,1:2], col="red")
lines(ebola_de[,1:2], col=rgb(0,0,0,0.7))
legend("topleft", 
       c("en", "de"), 
       col=c("red", rgb(0,0,0,0.7)),
       lty=1
       )
```

The similarities are striking. 


## Package specific time functions

Because data received from stad.grok.se is not always clean -- one might e.g. get dates like `2008-02-30` which is impossible -- the package has its own error robust date functions.

Furthermore, these functions work on all kinds of date formats like Date, numeric, character, POSIXlt, and POSIXct without having to make transformations all the time. The downside of this implementation is that edecuted guesses have to be made: 

  - character data is assumed to be given in format "yyyy-mm-dd" like in `r Sys.Date()`
  - numerics are assumed to be days since `1970-01-01` (which is R's default anyways)
  
To conclude, wikipediatrend time functions are easy to use efficient little helpers to work with the data provided by the package but are to be used with caution outside the package due the fact that convenience is based on educated guesses that can go wrong. 

1. `wp_date()`
  - is equivalent to `as.Date()` from the base package:
    - except that it will return `NA` in response to rediculous dates instead of an error
    - it will always assume `1970-01-01` to be the origin whenever numerics have to be tranfered to date but no origin is supplied 
2. `wp_day()`
  - extracts the day from a data
3. `wp_month()`
  - extracts the month from a date
4. `wp_month()`
  - extracts the year from a date
5. `wp_wday()`
  - extracts week day from a data (1 for Mondays, 2 for Tuesdays, ...)
6. `wp_yearmonth()`
  - extracts the year and month of a date and glues them together -- e.g. `2014-03-05` gets transformed to `"201403"`



# Credits

- Parts of the package's code have been shamelessly copied and modified from R base package written by R core team. This concerns the `wp_date()` generic and its methods and is detailed in the help files. 



<!-- http://www.tandfonline.com/doi/pdf/10.1080/10410236.2011.571759 -->



